Multi-resolution Neural Field Estimation

### 2011-03-08 Multiresolution_framework/ltx/manuscript.tex
some minor edits with Parham

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are the ideal candidates for estimation algorithms due to their strong links with the underlying physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=1-T_s/\tau$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an i.i.d. disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$: a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})$. A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function. 

@@

The MRAIDE can be implemented using various classes of scaling and wavelet functions. B-spline functions were chosen due their compact support and the ability to analytically calculate the convolution and inner product to form $\boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how $\boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$ are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 

@@

\section{Results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 200 realizations of 1~s of data were generated, where the estimation was applied to the final 900~ms, allowing the model's initial transients to die out. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance was set to $\eta(r-r') = \phi_{3,-2}(r-r')$ (B-spline, $j=3$). The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The firing rate slop, $\varsigma$ was set to $0.56~mV^{-1}$. The distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed.  %\ken{if we need to gain space I would remove the rest of this paragraph and put all parameters in a table} 
The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, \parham{This was obtained by averaging the power (over time \dean{realization?}) of the spatial frequency from the observations} \cite{Freestone2011}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

@@

The actual connectivity kernel and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}(a). \dean{No assumptions where made about the shape of the kernel.} \parham{It was assumed that the support of the kernel is unknown to the estimator.} The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator. \figurename{\ref{fig:KernelEstimate}}(b) and (c) shows the kernel scaling and wavelet functions, respectively.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig1.pdf}
		\caption{Actual (grey) and estimated (black) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 1.93$. (b) $j=1$, $n_x=33$, $RMSE = 1.39$. (c) $j=2$, $n_x=65$, $RMSE = 0.90$. (d) $j=3$, $n_x=131$, $RMSE = 0.78$. (e) $j=4$, $n_x=263$, $RMSE=0.78$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig2.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The mean kernel estimates over 200 realizations and \mike{confidence interval} are shown by the red line and red shaded region ($\pm2$ std.). (b) Kernel scaling functions at $j=1$. (c) Kernel wavelet functions at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution FEM representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

@@

\mike{The framework makes two key assumptions about the cortex: a linear activation function and stationary dynamics. Our main aim for this framework is to relax these assumptions using the multi-resolution decomposition. \dean{(our main wasn't really to relax the assumptions)} While the decomposition itself holds, efficiently performing nonlinear smoothing in the much larger state space remains a challenge. Additional avenues for future work include extending the model to capture a second order synaptic response kernel, and time delays at lower spatial resolutions.  The estimation techniques should also be extended to deal with spatial and temporal heterogeneity in the kernel. Finally, the majority of our current work involves applying this framework to collected data, primarily for the use of seizure detection and mitigation.}

>>>>

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics of the EEG. These models typically describe the macroscopic dynamics of the human brain, but are also descriptive of finer-scale neurodynamics. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

@@

where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=1-T_s/\tau$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the connectivity kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The term $e_t(\mathbf r)$ is an i.i.d. disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$: a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r})$. \dean{A linear form of the activation function was chosen to construct the estimator due to high computational demands when using a nonlinear (sigmoidal) activation function. }

@@

The MRAIDE can be implemented using various scaling and wavelet functions. B-spline functions were chosen due to their compact support and the ability to analytically define the convolution and inner product to form $\boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$. To show how $\boldsymbol\Lambda_x$ and $\boldsymbol \Lambda_{\theta}$ are formed, the $m^{th}$ order cardinal B-spline scaling function is defined by the recurrence relation \cite{Chui1992} 

@@

\section{Results}
To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically  using \eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}, enabling a comparison between true and estimated parameters. 200 realizations of 1~s of data were generated, where the estimation was applied to the final 900~ms, allowing the model's initial transients to die out. The observation noise was set to $\boldsymbol\Sigma_{\epsilon}=0.1 \times \mathbf{I}_{n_y}$ and the disturbance covariance was set to $\eta(r-r') = \phi_{3,-2}(r-r')$ (B-spline, $j=3$). The sampling period and the synaptic time constant were set to $T_s = 1$~ms and $\tau = 10$~ms, respectively. The firing rate slop, $\varsigma$ was set to $0.56~mV^{-1}$. The distance between adjacent sensors was $0.5$~mm, resulting into $n_y = 161$ observations. The observation kernel was modeled using a B-spline function, with a width of 0.8~mm at half the maximum amplitude. The spacing and bandwidth of the sensors allowed for the full spatial bandwidth of the field to be observed.  %\ken{if we need to gain space I would remove the rest of this paragraph and put all parameters in a table} 
The spatial cutoff frequency of the observed field, $y_t(r_n)$, was used to specify the level of decomposition required to represent the field using the wavelets. Using an oversampling parameter of 5, the cutoff spatial frequency was $\nu_{cy} \approx 6.9 $ cycles/mm, \parham{This was obtained by averaging the power in the spatial frequency (over time) of the spatial frequency from the observations} \cite{Freestone2011}. Therefore, wavelets up to level $j=3$ (with the bandwidth $\approx[5,8]$ cycles/mm) can represent the significant spatial characteristics of the field, yielding $n_x = 131$ states.

@@

The actual connectivity kernel and the decomposition is plotted in \figurename{\ref{fig:KernelEstimate}}(a). No assumptions where made about the shape of the kernel. The reconstructed kernel is in good accordance with the actual kernel, where the actual kernel lies inside the confidence interval. The small error in the estimate is attributed to the MRA of the system used to form the estimator. \figurename{\ref{fig:KernelEstimate}}(b) and (c) shows the kernel scaling and wavelet functions, respectively.
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig1.pdf}
		\caption{Actual (grey) and estimated (black) neural fields at a time instant for different spatial resolutions. (a) $j=0$, $n_x=17$, $RMSE = 1.93$. (b) $j=1$, $n_x=33$, $RMSE = 1.39$. (c) $j=2$, $n_x=65$, $RMSE = 0.90$. (d) $j=3$, $n_x=131$, $RMSE = 0.78$. (e) $j=4$, $n_x=263$, $RMSE=0.78$.}
	\label{fig:FieldEstimates}
\end{figure} 
\begin{figure}[!h] 
	\centering
		\includegraphics[scale=1]{./Graph/fig2.pdf}
		\caption{Connectivity kernel estimate. (a) The actual kernel is shown with the black solid line. The mean kernel estimates over 200 realizations and \mike{confidence interval} are shown by the red line and red shaded region ($\pm2$ std.). (b)-(c) Kernel scaling and wavelets functions, respectively, at $j=1$.}
	\label{fig:KernelEstimate}
\end{figure}
% \caption{The actual connectivity kernel is shown with the black solid line. The estimated kernel and confidence interval is shown by the red line and red shaded region ($\pm2$ std.). The 7 weighted kernel basis functions are shown by the black dashed lines.}
\section{Discussion}
In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution FEM representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling. For example, it is thought that the dynamics and connectivity structure differs at different spatial scales~\cite{Qubbaj2009}.

@@

\mike{The framework makes two key assumptions about the cortex: a linear activation function and stationary dynamics. Our main aim for this framework is to relax these assumptions using the multi-resolution decomposition. \dean{(our main aim wasn't really to relax the assumptions. if it was, we failed.)} While the decomposition itself holds, efficiently performing nonlinear smoothing in the much larger state space remains a challenge. Additional avenues for future work include extending the model to capture a second order synaptic response kernel, and time delays at lower spatial resolutions.  The estimation techniques should also be extended to deal with spatial and temporal heterogeneity in the kernel. Finally, the majority of our current work involves applying this framework to collected data, primarily for the use of seizure detection and mitigation.}
